A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-02-01 analysis/paper/paper.Rmd
update normalisation + synthetic data paper

Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Since 2014, already dozens of trajectory inference methods have been developed, each with its own set of methodological characteristics. This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking. We developed a framework of data structures for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available. In this work, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets. We employ a novel parameter optimisation procedure, and We compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness. We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.

@@

Where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:

$$
\begin{align}
P_j \propto \nu =  \rho \times \prod_j \left(\frac{y_j}{k_j}\right)^{n_j}
\end{align}
$$

@@

To each differential equation, we added an additional stochastic term:

$$
\begin{align}
\frac{dx_i}{dt} = m \times f(y_1, y_2, ...) - \lambda \times x_i &+ \eta \times \sqrt{x_i} \times \Delta W_t \\
\frac{dy_i}{dt} = r \times x_i - \Lambda \times y_i &+ \eta \times \sqrt{y_i} \times \Delta W_t
\end{align}
$$

@@

These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 2$.

@@

Estimation

Simulation

@@

DTW

## Common normalisation pipeline
We used a popular single-cell RNA-seq pipeline [@lun_stepbystepworkflowlowlevel_2016] which uses the scran and scater Bioconductor packages. The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed. Furthermore, the pipeline can select the most highly variable genes.
## Evaluation metrics
## Trajectory types

>>>>

Using single-cell transcriptomics data, it is now possible to computationally order cells along trajectories, allowing the unbiased transcriptome-wide study of cellular dynamic processes. Since 2014, more than 50 trajectory inference methods have been developed, each with its own set of methodological characteristics. This makes a direct comparison between trajectory inference methods often difficult, and as a result a comprehensive assessment of the performance and robustness of each method is still lacking. We developed a framework of data structures for trajectory inference methods in order to make their results comparable to each other and to a gold standard, when available. In this work, we compare the results from a total of 22 trajectory inference methods, on a large collection of real datasets and realistic synthetic datasets. We employ a novel parameter optimisation procedure, and We compare methods using several metrics, including accuracy of the inferred ordering, correctness of the network wiring, code quality and user friendliness. We conclude with practical guidelines for method users, and also believe our evaluation framework can be used to spearhead the development for new methods. Our evaluation pipeline is fully reproducible, can be easily extended, and is available at https://www.github.com/dynverse.

@@

Where the hill coefficient $n$ represents the cooperativity of binding and $k$ the transcription factor concentration at half-maximal binding. When multiple regulators are bound:
$$
\begin{align}
P_j \propto \nu =  \rho \times \prod_j \left(\frac{y_j}{k_j}\right)^{n_j}
\end{align}
$$

@@

To each differential equation, we added an additional stochastic term: 
$$
\begin{align}
\frac{dx_i}{dt} = m \times f(y_1, y_2, ...) - \lambda \times x_i &+ \eta \times \sqrt{x_i} \times \Delta W_t \\
\frac{dy_i}{dt} = r \times x_i - \Lambda \times y_i &+ \eta \times \sqrt{y_i} \times \Delta W_t
\end{align}
$$

@@

These SDEs were simulated using the Euler–Maruyama approximation, with time-step $h = 0.01$ and noise strength $\eta = 8$. The total simulation time varied between 5 for linear and bifurcating datasets, 10 for consecutive bifurcating, trifurcating and converging datasets, 15 for bifurcating converging datasets and 30 for linear long, cycle and bifurcating loop datasets. The burn-in period was for each simulation 2. Each network was simulated 32 times.

@@

For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in. Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities. We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$. These were combined with the expression levels of the genes simulated within a trajectory. Next, true counts were simulated using $Y'_{i,j} \sim \text{Poi}(\lambda_{i,j})$. Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernouilli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.

@@

Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory structure is not straightforward. Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable sfig?. To estimate a cell's position during a simulation within the trajectory structure, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone. We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation. We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory structures, the number of possible milestones a backbone could progress through was limited to 20.

## Common normalisation pipeline
We used a standard single-cell RNA-seq preprocessing pipeline which uses the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016. The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3). We required that a gene has to be expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05. Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5. As a final filter, we removed both all zero genes and all zero cells until convergence.
## Evaluation metrics
## Trajectory types
